Hydra Battles and Cie

### 2021-01-19 doc/chapter-primrec.tex
redaction of Ackermann chapter (continued)

\vspace{4pt}
\noindent
\emph{From\href{../theories/html/hydras.Ackermann.PrimRecExamples.html}{Ackermann.PrimRecExamples}}.
\begin{Coqsrc}
Lemma  const1_NIsPR n : isPR 0 n. 
Proof.
  induction n.
  - apply zeroIsPR.
  - destruct IHn as [x Hx].
   exists (composeFunc 0 1 (PRcons 0 0 x (PRnil 0)) succFunc). 
   cbn in *; intros; now rewrite Hx.
Qed.
\end{Coqsrc}

@@

@@

\begin{itemize}
\item Write a simple and readable proof that the Fibonacci function is primitive recursive.

@@

Definition fib_tail (n:nat) := general_fib 1 1 n.
\end{Coqsrc}

\item Prove formally that the Ackermann function is \emph{not} primitive recursive.

\end{itemize}

>>>>

\vspace{4pt}
\noindent
\emph{From \href{../theories/html/hydras.Ackermann.PrimRecExamples.html}{Ackermann.PrimRecExamples}}.
\begin{Coqsrc}
Lemma  const1_NIsPR n : isPR 0 n. 
Proof.
  induction n.
  - apply zeroIsPR.
  - destruct IHn as [x Hx].
   exists (composeFunc 0 1 (PRcons 0 0 x (PRnil 0)) succFunc). 
   cbn in *; intros; now rewrite Hx.
Qed.
\end{Coqsrc}

\subsubsection{Proving that \texttt{plus} is primitive recursive}

The lemma \texttt{plusIsPR} is already proven in \href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}. We present in 
\href{../theories/html/hydras.Ackermann.PrimRecExamples.html}{Ackermann.PrimRecExamples}
a commented version of this proof,

First, we look for lemmas which may help to prove that a given function obtained with the recursor \texttt{nat\_rec} is primitive recursive.

\begin{Coqsrc}
Search (is_PR 2 (fun _ _ => nat_rec _ _ _ _)).
\end{Coqsrc}

\begin{Coqanswer}
ind1ParamIsPR:
  forall f : nat -> nat -> nat -> nat,
  isPR 3 f ->
  forall g : nat -> nat,
  isPR 1 g ->
  isPR 2
    (fun a b : nat =>
     nat_rec (fun _ : nat => nat)
                 (g b) (fun x y : nat => f x y b) a)
\end{Coqanswer}

We prove that the library function \texttt{plus} is extensionally equal to a function defined with
\texttt{nat\_rec}.

\begin{Coqsrc}
Definition plus_alt x y  :=
              nat_rec  (fun n : nat => nat)
                       y
                       (fun z t =>  S t)
                       x.

Lemma plus_alt_ok:
  extEqual 2 plus_alt plus.
Proof.
  intro x; induction x; cbn; auto.
  intros y; cbn; now rewrite <- (IHx y).
Qed.
\end{Coqsrc}

A last lemma before the proof:

\begin{Coqsrc}
Lemma isPR_extEqual_trans n : forall f g, isPR n f ->
                                    extEqual n f g ->
                                    isPR n g.
Proof.
 intros f g [x Hx]; exists x.
 apply extEqualTrans with f; auto.
Qed.
\end{Coqsrc}

Let us start now.

\begin{Coqsrc}
Lemma plusIsPR : isPR 2 plus.
Proof.
  apply isPR_extEqual_trans with plus_alt.
  - unfold plus_alt; apply ind1ParamIsPR.
\end{Coqsrc}

\begin{Coqanswer}
2 subgoals (ID 126)

  ============================
  isPR 3 (fun _ y _ : nat => S y)

subgoal 2 (ID 127) is:
 isPR 1 (fun b : nat => b)
\end{Coqanswer}

We alreday proved that \texttt{S} is \texttt{PR 1}, but we need to consider a function of three arguments, which ignores its first and third arguments.
Fortunately, the library \texttt{primRec} already contains lemmas adapted to this kind of situation.

\begin{Coqanswer}
filter010IsPR :
forall g : nat -> nat, isPR 1 g -> isPR 3 (fun _ b _ : nat => g b)
\end{Coqanswer}

Thus, our first subgoal is solved easily and the proof ends, applying alreday proven lemmas.

\begin{Coqsrc}
 - unfold plus_alt; apply ind1ParamIsPR.
    + apply filter010IsPR, succIsPR.
    + apply idIsPR.
  - apply plus_alt_ok. 
Qed.
\end{Coqsrc}

@@

\begin{exercise}
There is a lot of lemmas similar to \texttt{filter010IsPR} in the \texttt{primRec} library, useful to control the arity of functions.
Thus, the reader may look at them, and invent simple examples of application for each one.
\end{exercise}

\begin{exercise}
Multiplication of natural number is already proven in the \texttt{primRec} library. Write a proof of your own, then compare to the library's version.
\end{exercise}

\subsubsection{More examples}

The following proof decomposes the \texttt{double} function as the composition of 
multiplication,with the pair composed of the identity and the constant function wihich returns $2$.
\emph{Note that the lemma \texttt{const1\_NIsPR} considers this function as an unary function (unlike \texttt{const0\_NIsPR} }.

\begin{Coqsrc}
Lemma doubleIsPR : isPR 1 double.
Proof.
  unfold double; apply compose1_2IsPR.
  - apply idIsPR.
\end{Coqsrc}

\begin{Coqanswer}
subgoal 1 (ID 110) is:
 isPR 1 (fun _ : nat => 2)
subgoal 2 (ID 111) is:
 isPR 2 Init.Nat.mul
\end{Coqanswer}

 \begin{Coqsrc}
  - apply const1_NIsPR.
  - apply multIsPR.
Qed.
\end{Coqsrc}

\begin{exercise}
Consider the following functions:

\begin{Coqsrc}
Fixpoint fac n :=
  match n with 
          | 0 => 1
          | S p  => n * fac p
  end.

Fixpoint exp n p :=
  match p with
  | 0 => 1
  | S m =>  exp n m * n
  end.

Fixpoint tower2 n :=
  match n with
  | 0 => 1
  | S p => exp 2 (tower2 p)
  end.
\end{Coqsrc}

Prove they are primitive recursive.

\textbf{Hint:} You may have to look again at the lemmas of the library
\href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec} if you meet some difficulty.
\end{exercise}

@@

The following projects may require to look at other libraries, and or prove new lemmas.
Please send us your solutions (even if they are partial).

\index{Projects}

\begin{project}
Write a simple and readable proof that the Fibonacci function is primitive recursive.

@@

Definition fib_tail (n:nat) := general_fib 1 1 n.
\end{Coqsrc}
\end{project}

\begin{project}
Prove formally that the Ackermann function is \emph{not} primitive recursive.

You may find a definition and some useful lemmas in
 \href{../theories/ordinals/Prelude/Iterates.html}%
{ordinals.Prelude.Iterates}.

\end{project}
